---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# USAID Flood Response Rehabilitation and Chlorination Survey

<!-- badges: start -->
[![R-CMD-check](https://github.com/openwashdata/floodchlorinationsurvey/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/openwashdata/floodchlorinationsurvey/actions/workflows/R-CMD-check.yaml)
[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15846578.svg)](https://zenodo.org/doi/10.5281/zenodo.15846578)
<!-- badges: end -->

This dataset originates from the USAID Flood Response Rehabilitation and
Chlorination Survey, conducted between 2019 and 2020 in Mulanje
District, Malawi. The data was collected using the mWater mobile
platform by field teams working to assess, rehabilitate, and chlorinate
rural handpump-equipped water points affected by flooding.

The dataset includes detailed records on:

-   Borehole component condition before and after rehabilitation
-   Chlorination activities
-   Community mobilization
-   Environmental risks near water points
-   Functionality status and water access metrics post-rehabilitation

Photos and GPS coordinates are embedded via cloud-linked URLs to provide
visual and spatial evidence of the rehabilitation process.

**Potential Use Cases**

1.  WASH Program Monitoring and Evaluation

-   Track the impact of rehabilitation efforts on water point
    functionality.

-   Monitor the effectiveness of chlorination in flood-prone regions.

2.  Infrastructure Asset Management

-   Identify components frequently requiring replacement.

-   Schedule preventive maintenance and prioritize resources.

3.  Disaster Response Planning

-   Map flood-affected water infrastructure and assess repair timelines.

-   Use historical damage and repair data to prepare for future climate
    shocks.

4.  Public Health Risk Assessment

-   Analyze the relationship between water source integrity and
    contamination risks (e.g., proximity to latrines or cemeteries).

5.  Policy and Program Design

-   Guide government and donor decisions on where and how to invest in
    resilient rural water systems.

6.  Training and Capacity Building

-   Use the data and photos as case studies for training technicians or
    water committees.

Potential Users of this data

1. Government agencies (e.g., Ministry of Water, District Water
Offices)

2.  NGOs and implementing partners involved in WASH and emergency
    response

3.  Donor agencies such as USAID, to assess project outcomes and guide
    future funding

4.  Researchers and public health analysts studying the impacts of
    infrastructure damage and recovery

5.  Water point mechanics and private sector suppliers, to inform
    maintenance and procurement planning

6.  Community-based organizations (CBOs) and Water Point Committees, for
    local monitoring and advocacy

## Installation

You can install the development version of floodchlorinationsurvey from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/floodchlorinationsurvey")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/floodchlorinationsurvey/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to the USAID Flood Response Rehabilitation and Chlorination Survey, conducted between 2019 and 2020 in Mulanje District, Malawi.

```{r}
library(floodchlorinationsurvey)
```

### floodchlorinationsurvey

The dataset `floodchlorinationsurvey` contains 
`r nrow(floodchlorinationsurvey)` observations and
`r ncol(floodchlorinationsurvey)` variables

```{r}
floodchlorinationsurvey |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "floodchlorinationsurvey.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
library(floodchlorinationsurvey)

# Visualization: Most Frequently Replaced Components
# Purpose:  Highlight common mechanical failures. 

# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)

# Define the replacement tick variables (TRUE/FALSE indicating part was replaced)
tick_vars <- c(
  "replaced_rods_tick", "replaced_flapper_tick", "replaced_handle_tick",
  "replaced_cylinder_tick", "replaced_socket_tick", "replaced_rope_tick",
  "replaced_footvalve_bobbin_tick", "replaced_footvalve_body_tick",
  "replaced_plunger_bobbin_tick", "replaced_plunger_body_tick",
  "replaced_rodhanger_tick", "replaced_head_bolt_tick",
  "replaced_compression_cone_tick", "replaced_pump_head_case_tick",
  "replaced_pump_head_cover_tick", "replaced_rising_main_tick",
  "replaced_fulcrum_pin_tick", "replaced_fulcrum_bearings_tick",
  "replaced_hanger_pin_tick", "replaced_hanger_bearings_tick",
  "replaced_steel_cone_tick", "replaced_top_sleeve_tick",
  "replaced_rod_centralisers_tick", "replaced_rising_centralisers_tick"
)

# Count how many times each component was marked as replaced (TRUE)
component_freq <- floodchlorinationsurvey %>%
  # Select only the replacement tick columns
  select(all_of(tick_vars)) %>%
  # Summarise to count how many TRUEs per column
  summarise(across(everything(), ~ sum(. == TRUE, na.rm = TRUE))) %>%
  # Reshape the summary to long format: component | count
  pivot_longer(cols = everything(), 
               names_to = "component", 
               values_to = "count") %>%
  # Sort in descending order by count
  arrange(desc(count))

# Clean component names for better readability
component_freq$component <- gsub("replaced_|_tick", "", component_freq$component)  # Remove prefixes/suffixes
component_freq$component <- gsub("_", " ", component_freq$component)               # Replace underscores with spaces
component_freq$component <- tools::toTitleCase(component_freq$component)           # Capitalize each word

# Create a horizontal bar chart
ggplot(component_freq, aes(x = count, y = reorder(component, count))) +
  geom_bar(stat = "identity", fill = "#2a9d8f") +  
  labs(
    title = "Most Frequently Replaced Handpump Components",
    x = "Number of Replacements",
    y = "Component"
  ) +
  theme_minimal(base_size = 13) 
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/floodchlorinationsurvey/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("floodchlorinationsurvey")
```
